\textbf{ANALYSIS OF \\ DEEP REINFORCEMENT \\ APPLIED TO A BOARD GAME}

### 2023-02-16 doc/paper.tex

\begin{abstract}
    In October 2022 DeepMind announced an artificial intelligence in a preprint \cite{alphatensor} that was called AlphaTensor. It utilized \emph{artificial neural networks} and \emph{reinforcement learning} in order to solve the unsolved problem of finding more efficient algorithms for matrix multiplication. AlphaTensor was succesful and illustrated the impact of \emph{artificial neural networks} combined with \emph{reinforcement learning}. This paper takes a closer look into the effectiveness of \emph{artificial neural networks} and \emph{reinforcement learning} strategies such as the ones used in AlphaTensor. More precisely, the effects the parameters \emph{kernel size}, \emph{learning rate}, \emph{discount factor} and \emph{noise} have on the networks' performance in the board game \emph{Connect Four} are analyzed. Furthermore the results show whether the parameters have small or great impact on the performance.

@@

The concept of artificial intelligence was only popularized during the first half of the 20th century through science fiction. Computers before 1950 lacked key prerequisite for intelligence: they could not store commands, only execute them immediately. Allen Newell, Cliff Shaw and Herbert Simon created in 1956 what is often considered the first artificial intelligence program and it catalyzed the next 20 years of research in this field.

@@

\noindent
Gordon Moore predicted in 1965 that the memory and speed of computers would double every year and the prediction, known as Moore's law, has been true ever since. The growth led to new possibilities for the field of artificial intelligence and many landmark goals were achieved during the 1990's and 2000's. In 1997 reigning world chess champion and grandmaster Gary Kasparov was defeated by IBM's computer program Deep Blue \cite{deep_blue}. 19 years later Deepmind's artificial intelligence AlphaGo \cite{alphago} beat the South Korean grandmaster Lee Sedol regarded as one of the best Go players of all time. Go is due to its large search space considerably more complex than chess and AlphaGo taught itself how to play without any proprietary knowledge in comparison to Deep Blue which was heavily influensed by pre-coded strategies. Shortly after the success of AlphaGo Deepmind began working on protein structure prediction. The artificial intelligence AlphaFold \cite{alphafold} is now by a huge margine the most accurate in predicting protein structures and it is accelerating research in nearly every field of biology. As the field of aritificial intelligence visibly is becoming a substantial part of today's science we find it important to analyze different aspects of it.

@@

Artificial neural networks \cite{ann} are computing systems consisting of connected nodes called artifical neurons and are inspired by biological neural networks that constitute human brains. The artificial networks have recently been used in different forms in order to solve various complex and abstract problems. A recent example is Deepmind's AlphaTensor \cite{alphatensor} which in October 2022 found a more efficient algorithm for matrix multiplication.

@@

\noindent
The greatest factors of an artificial neural network's success is its architecture and how it is applied to the problem presented. AlphaZero \cite{alphazero} and AlphaGo \cite{alphago} are computer programs whose artifical neural networks only differ in the method of application. Even though AlphaZero uses the same type of network it beats AlphaGo 100 to 0 in the board game Go which demonstrates how delicate artificial networks can be. In this project we take a closer look at the intricacies of artificial neural networks with the help of the board game Connect Four. We have chosen to study the following aspects: Kernel Size, Learning Rate and Training Noise. The aspects are described further in later sections.

@@

Following are the questions researched in this paper. The details of the questions are going to be explained in later sections.

@@

This project is limited to a less complex game due to the limitations of computing power and time. Board games such as Chess or Shogi have large search spaces due to the number of different moves possible from each board state and are therefore not suitable for this project. A board game such as Tic-tac-toe is on the other hand not complex enough to analyze features such as the kernel size of a convolutional neural network. We therefore chose the board game Connect Four which does not have a too large search space and is complex enough where convolutional kernels can have different effects.

@@

No time will be put into making our Connect Four board user-friendly since it is not the purpose of this project. Time will on the other hand be spent making the board and its operations efficient in order to make training and playing consume less time.

@@

The goal of this project is neither to create a perfect artificial intelligence nor a perfect player. It is rather to analyze what parameters play a role in making an artificial neural network better and what makes it worse. Deficiencies are even welcome as they provide stronger insight in what to avoid when creating and training a neural network.

@@

The purpose of this paper is to inform readers with no proprietary knowledge about machine learning. It is therefore necessary that we concisely present all of the topics from the beginning. Giving a deeper understanding of the theory is a bigger priority for us than presenting formal mathematical descriptions of everything which could end up making the readers more confused.

@@

The nature of learning is often associated with the foundational idea of learning by interacting with the surrounding environment. When a child learns how to ride a bicycle, the child either falls or manages to stay in balance. Everytime parents speak to their children they indirectly correct the children's sentences by repeating them correctly. Constantly interacting with the environment produces a wealth of information about consequences of actions, about the cause and effect of them and about what actions to take in order to achieve a certain goal. These kinds of interactions undoubtedly play a major role throughout our lives as we learn about ourselves and the environment surrounding us. Whether it is learning how to ride a bike or how to speak, we are always aware of how the environment responds to our actions. By behaving in certain manner and choosing actions carefully, we can reach the goals we seek. Reinforcement learning \cite{sutton} offers a computational approch to learning from interaction. By primarily exploring idealized learning situations and evaluting the effectiveness of various learning methods, reinforcement learning focuses a lot more on goal-oriented learning from interaction than any other machine learning approach.

@@

Following are examples in order to better grasp the concept of reinforcement learning and how it could be applied to real life problems.

@@

\begin{outline}
    \1 Adam left his house five minutes ago and is on his way to school. He suddenly starts questioning whether he locked the front door or not. Turning home would at this point of time make him late to class but leaving the house unlocked has its risks. Adam makes his decision based on the time he gets home after school and how often he has actually forgotten to lock the door. 
    \1 Hugo is playing chess against the current chess world champion Magnus Carlsen. In order to not lose, Hugo makes moves by intuitive judgement of the current position and by anticipating possible counterreplies. Since Hugo is familiar with the current position and has played it at home several times, he is quite confident in his decisions. 
    \1 Romeo could hardly move his body the first few months after being born. One year later he was running at 1.6 meters per second.
    \1 Jonathan is in a casino and suddenly sees 10 slot machines. He does not know the probability of hitting a jackpot on any of the machines but knows the cost of each lever pull and the reward given by a jackpot. Through repeated selections of slot machines, Jonathan manages to maximize his winnings by concentrating on the best levers. 
\end{outline}

@@

\noindent
All of the examples share features that are easily overlooked due to their simplicity. They all involve interaction between a decision-making agent and an environmnet, where the agent is trying to achieve a particular goal without knowing the full extent of the environment nor the consequences of its actions. The agent's actions are able to affect the environment (e.g., Hugo making a move against Magnus Carlsen changes the state of board) thereby affecting the actions available to the agent in the future. None of the actions, in all of the examples, can be predicted fully and the agent must instead observe the environment and how it changes based on actions in order to learn how to properly behave.

@@

As previously illustrated there are two main components of reinforcement learning, an agent and an environment. It is on the other hand possible to recognize three more elements of a reinforcement learning system. Namely a policy, a reward signal and a value function.

@@

\noindent
Further, the \emph{reward signal} \cite{sutton} determines the goal of the learning agent. After each decision the agent makes, the environment sends a reward signal and the goal is to maximize the total reward after all of the actions have been made. The reward signal thus determine what actions are and bad respectively. This could be compared to pain and pleasure in human beings. When a child falls of its bike it recieves a negative reward signal in the form of pain. On the other hand, maintaining balance on the bike for a certain amount of time sends a positive reward signal in the form of happiness. Since the child, the learning agent, wants to experience joy instead of pain the goal becomes riding a bike without falling off of it.

@@

A Markov Decision Process \cite{sutton}, MDP, is a more rigorous and mathematical way of describing a seqentual decision-making system. A MDP can formally be defined as MDP$(S, A, R)$ where each parameter is defined as following:

@@

\noindent
More specifically, an agent and an environment interact with eachother in a MDP at discrete time steps, $t = 0, 1, 2, 3, ...$. The agent recieves a representation of the environment's state, $S_{t} \in S$, at each time step and accordingly selects an action, $A_{t} \in A(S)$. The time step following, $t + 1$, as a consequence of its action, the agent recieves a reward signal, $R_{t} \in R$, and arrives at a new state, $S_{t+1}$. The MDP thereby gives rise to following sequence:

@@

\noindent
In our case with Connect Four, the board becomes the environment and the players are agents taking turn to interract with it in the Markov Decision Process.

@@

\vskip 0.2cm
In 1989 Watkins \cite{sutton} released one of the early breakthroughs in reinforcement learning. It was the algorithm known as Q-learning which is a way for an agent to learn how to optimally navigate in a Markov Decision Process (3.4). The Q-learning algorithm is an action-value function which maps actions to values, where the action-values are called Q-values. The Q-learning algorithm evaluates how benificial a certain action is in a given situation and a higher Q-value corresponds to a better action. The evalutations made by the algorithm improve over time. The algorithm \cite{sutton} is formally defined as follows:

@@

\noindent
\emph{Gamma}, $\gamma$, is in this expression known as the discount factor and it changes the behaviour in the agent's actions. Values closer to 0 lead to the reward $R_{t}$ propagating worse from the state $S_{t}$ to $S_{0}$ which encourages the agent to take actions that yield immediate rewards. A discount factor closer to the value 1 will on the other hand encourage the agent to take actions that yield later rewards since a larger portion of the reward propogates through the states in this case.

@@

\noindent
\emph{Alpha}, $\alpha$, is in this expression known as the learning rate and regulates the agent's rate of learning, just as the name suggests. Values closer to 0 will decrease the update in the formula above which decreases the rate of learning. A learning rate closer to 1 will on the other hand increase the rate of learning as a large part of the update value is kept.

@@

Since the future maximum action value, $\max_{a}Q(S_{t+1}, a)$, in Q-learning is evaluated using the same Q functions as the one used by the agent to choose the action, $a_{t}$, the algorithm risks overestimating certain action values \cite{doubleq}. In our case with Connect Four, it could be evaluating bad moves that randomly led to a victory the same as good moves that actually lead to a victory. This slows down the learning due to the fact that the Q-learning algorithm will later need to re-adjust the overestimated evalutations. A variant called double Q-learning \cite{doubleq} was proposed to correct this. In practice, it uses two seperate value functions, $Q_{X}$ and $Q_{Y}$, that are updated in a mutually symmetric fashion using eachother. Double Q-learning is defined as follows:

@@

\noindent
Now that the future maximum action is evaluated using a seperate value function the overestimation issue has been solved. The double Q-learning algorithm could be seen as two people checking that each other's homework is done correctly which naturally increases the correctness and in turn the effectiveness.

@@

Artificial neural networks \cite{charu} are simulations of the mechanism of learning in biological organisms. The human brain and nervous system contain cells known as neurons. The neurons are connected to eachother with the help of synaptic connections. The strength of synaptic connections change in response to external stimuli and that is how learning takes place in not only humans but also many other living organisms.

@@

\noindent
This learning is simulated in artificial neural networks \cite{charu} which contain artificial neurons. The neurons are connected to eachother through weights that immitate the strength of synaptic connections in living organisms.

@@

\noindent
The basic architecture of the perceptron is shown in Figure 10 above. The input layer consists of input nodes that each are connected to the output node through weights, $w_{1}, w_{2}, ..., w_{6}$. The input nodes do not perform any changes to the input values, $x_{1}, x_{2}, ..., x_{6}$ and send the input values to the output node through the corresponding weight connections. The output node then recieves the sum of each input value multiplied by the weight it was sent through, a weighted sum. The node also contains an internal value called \emph{bias}, $b$, that is added to the sum. If we let $v$ denote the output node's recieved value, then $v = b + \sum w_{i} x_{i} = w_{1} x_{1} + w_{2} x_{2} + ... + w_{6} x_{6}$. Before the output then sends the value further, it applies an \emph{activation function} which often is non-linear. If we let $f$ denote the activation function, the output of the output node becomes $y = f(v)$. This procedure where a set of input values, $x_{i}$ are transformed into a set of output values, $y_{i}$, by the neural network is called a \emph{forward pass} \cite{charu}.

@@

Multilayer neural networks consist of more than one computational layer \cite{charu} in contrast to the perceptron. The perceptron only consists of an input and output layer, where the output layer is the only layer that performs computations. Multilayer neural networks on the other hand contain intermediate layers in addition to the input and output layer. The intermediate layers are between the input and output layer and are often referred to as \emph{hidden layers} since the computations performed by them are not visible. The weight connections, internal biases and the mechanics of the forward pass remain the same as in the perceptron, with the only difference being that there are several nodes in each hidden layer that recieve a weighted sum of values from the preceding layer.

@@

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.6]{multilayered}
    \vskip 0.2cm
    \caption{The basic architecture of a multilayerd neural network with two hidden layers and a single output layer. Note that the input layer is not included when we count the number of layers since it only transmits the input data into the network. }
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.3]{feedforward}
    \caption{A forward pass in a multilayer neural network with 1 hidden layer and 1 output layer. Weighted sums from the preceding layers are in orange. The values in each node represents its internal bias.}
\end{figure}

\noindent
Figure 12 above illustrates a multilayer neural network with 2 input nodes, followed by a hidden layer with 3 nodes which connencts to a single output node. Both the hidden and the output layer use a linear activation function, $f(x) = x$, for simplicity. The two input values are each 1 and the internal bias, $B$, of input nodes are always 0. Let each node be indexed as $(i, j)$ where $i$ is the index of the layer from left to right, zero-indexed, and $j$ is the index of the node in its layer from top to bottom, zero-indexed. Furthermore let $W_{(i_{1}, j_{1}) (i_{2}, j_{2})}$ denote the weight connection between node $(i_{1}, j_{1})$ and $(i_{2}, j_{2})$. The value, $V$, of each node and the final output, $y$, are then calculated as follows:

@@

Figure 12 shows a neural network that only uses a linear activation function, $f(x) = x$, but it is essential that non-linearity is introduced into the network. Non-linearity means that the output of the neural network cannot be reproduced from a linear combination of the input values. Without a non-linear activation function in the neural network, no matter how many layers it has, it would behave like a simple perceptron, since combining the layers would only produce another linear function. By introducing non-linear activation functions into the neural network, it is able to approximate models that vary non-linearly in addition to linear models. The activation function used throughout this project is the hyperbolic tangent function, \emph{tanh}, which has the output range $[-1, 1]$ and is defined as follows:

@@

A neural network can be interpreted as a large statistical model with a large amount of parameters, weight connections. If each of the parameters, weight connections, were to be tuned well, then the model would be able to represent highly abstract and complex correlations. A less rigorours way of imagining this would be having a function in a space with as many dimensions as weight connections in the network. The phenomena that that function would be able to approximate would be beyond our understanding. This leads to the most common usage of neural networks. Namely training neural networks by adjusting their weight connections in order for it to approximate a certain function with the help of data. An example would be approximating the sinus curve with a neural network. By giving an x-coordinate as input to the network and comparing the networks output value with the y-coordinate one only needs to adjust the weight connections of the network in such a manner that the error decreases in order to make it a better approximation. This is further described in following sections.

@@

To be able to train a neural network to certain level we first of all need to know at what level it currently is at. An efficient method to measure this is to find an error. Given a output target and the neural network's output, the error can be described as a function of them. In order to then improve the network we simply need to find a way to decrease that error. This type of function which is to be minimized is also referred to as a \emph{loss function} \cite{charu}. How the loss function is used in order to train a neural network is described in later sections. The loss function used throughout this project is the \emph{Mean Squared Error} \cite{charu}, MSE, and is defined over a set of network predictions or outputs, $\hat{y_{1}}, \hat{y_{2}}, ..., \hat{y_{n}}$, and output targets, $y_{1}, y_{2}, ..., y_{n}$ as follows:

@@

\noindent
An advantage of The MSE is that it ensures that our trained will not have outlier predictions with prominent errors, since the MSE puts larger weight on those errors due to the squaring in the function. A disagvantage is on the other hand that a single very poor prediction made by the network will overly impact the training since the error is magnified by the squaring part of the MSE function. Since we most of the cases aim for more of a well-rounded neural network that performs well on most of the data those outliers should be ignored and not prioritized. This is not a big problem in our project since our networks only output values in the range $[-1, 1]$ which ensures that the squaring in the MSE does not magnify any errors by a lot.

@@

Assume we are trying to train a perceptron with exactly one input node and one output node with one weight connection, $w$, between them. For the sake of simplicity let the output node's activation function be linear, $f(x) = x$. Let the input value be a constant $x$, the output target be a constant $y$ and the output node's internal bias be a constant $b$. The error is then defined as follows:

@@

\noindent
Notice that $w$ is the only variable in the error and is exactly what needs to be adjusted as mentioned in section 4.3. If we now plot the MSE function as a function of $w$ things become more apparent. The second degree polynomial created by the MSE in respect to the weight, $w$, is illustraded below, non-scaled:

@@

\noindent
As Figure 15 shows the local minimum of the MSE and finding a weight $w$ such that the loss function is at the local minimum is equivalent to minimizing the error, which is desired as described in section 4.4.1. One way of finding that local minimum is to descend the loss function by using its slope at the current weight, also known as the derivative in that point. Note that we are assuming that the loss function is defined and differentiable around that point. Since the slope describes the 'steepness' of the loss function in the current point, following the direction of that slope will either lead you to a local maximum or minimum. It could be thought of as walking up or down a hill as illustrated by Figure 15 above. The slope of the MSE changes as one walks down the slope, assuring that the movement is towards the minimum. This type of descent will be described more rigorously in the following sections.

@@

In order to rigorously generalize the descention algorithm described in section 4.4.2 a new mathematical tool needs to be introduced. Namely the \emph{gradient} of a function. The example used in section 4.4.2 was a perceptron with a single output node and a single input node. Having a multilayer neural network with several weight connectieons instead leads to the loss function being dependant on several variables, namely the weights $w_{1}, w_{2}, ..., w_{n}$. In order to find the slope of a function $f(w_{1}, w_{2}, ..., w_{n})$ in a point we employ a mathematical technique called \emph{partial derivatives}. Let us use a small example. Take the following simple case of $f$:

@@

\noindent
Using the expression in Figure 21 in combination with the expressino in Figure 20 the following recursion is attained:

>>>>

\begin{abstract}
    In October 2022 DeepMind announced an artificial intelligence in a preprint \cite{alphatensor} that was called AlphaTensor. It utilized \emph{artificial neural networks} and \emph{reinforcement learning} in order to solve the unsolved problem of finding more efficient algorithms for matrix multiplication. AlphaTensor was successful and illustrated the impact of \emph{artificial neural networks} combined with \emph{reinforcement learning}. This paper takes a closer look into the effectiveness of \emph{artificial neural networks} and \emph{reinforcement learning} strategies such as the ones used in AlphaTensor. More precisely, the effects the parameters \emph{kernel size}, \emph{learning rate}, \emph{discount factor} and \emph{noise} have on the networks' performance in the board game \emph{Connect Four} are analyzed. Furthermore the results show whether the parameters have a small or great impact on the performance.

@@

The concept of artificial intelligence was only popularized during the first half of the 20th century through science fiction. Computers before 1950 lacked key prerequisites for intelligence: they could not store commands, only execute them immediately. In 1956 Allen Newell, Cliff Shaw and Herbert Simon created what is often considered the first artificial intelligence program and it catalyzed the next 20 years of research in this field.

@@

\noindent
Gordon Moore predicted in 1965 that the memory and speed of computers would double every year and the prediction, known as Moore's law, has been true ever since. The growth led to new possibilities for the field of artificial intelligence and many landmark goals were achieved during the 1990's and 2000's. In 1997 reigning world chess champion and grandmaster Gary Kasparov was defeated by IBM's computer program Deep Blue \cite{deep_blue}. 19 years later Deepmind's artificial intelligence AlphaGo \cite{alphago} beat the South Korean grandmaster Lee Sedol regarded as one of the best Go players of all time. Go is due to its large search space considerably more complex than chess and AlphaGo taught itself how to play without any prior knowledge in comparison to Deep Blue which was heavily influenced by pre-coded strategies. Shortly after the success of AlphaGo Deepmind began working on protein structure prediction. The artificial intelligence AlphaFold \cite{alphafold} is now by a huge margin the most accurate in predicting protein structures and it is accelerating research in nearly every field of biology. As the field of artificial intelligence visibly is becoming a substantial part of today's science we find it important to analyze different aspects of it.

@@

Artificial neural networks \cite{ann} are computing systems consisting of connected nodes called artificial neurons and are inspired by biological neural networks that constitute human brains. Artificial networks have recently been used in different forms to solve various complex and abstract problems. A recent example is Deepmind's AlphaTensor \cite{alphatensor} which in October 2022 found a more efficient algorithm for matrix multiplication.

@@

\noindent
The greatest factors of an artificial neural network's success are its architecture and how it is applied to the problem presented. AlphaZero \cite{alphazero} and AlphaGo \cite{alphago} are computer programs whose artificial neural networks only differ in the method of application. Even though AlphaZero uses the same type of network it beats AlphaGo 100 to 0 in the board game Go which demonstrates how delicate artificial networks can be. In this project, we take a closer look at the intricacies of artificial neural networks with the help of the board game Connect Four. We have chosen to study the following aspects: Kernel Size, Learning Rate and Training Noise. The aspects are described further in later sections.

@@

The following are the questions researched in this paper. The details of the questions are going to be explained in later sections.

@@

This project is limited to a less complex game due to the limitations of computing power and time. Board games such as Chess or Shogi have large search spaces due to the number of different moves possible from each board state and are therefore not suitable for this project. A board game such as Tic-tac-toe is on the other hand not complex enough to analyze features such as the kernel size of a convolutional neural network. Therefore we chose the board game Connect Four which does not have a too large search space and is complex enough where convolutional kernels can have different effects.

@@

No time will be put into making our Connect Four board user-friendly since it is not the purpose of this project. Time will, on the other hand, be spent making the board and its operations efficient to make training and playing consume less time.

@@

The goal of this project is neither to create a perfect artificial intelligence nor a perfect player. It is rather to analyze what parameters play a role in making an artificial neural network better and what makes it worse. Deficiencies are even welcome as they provide stronger insight into what to avoid when creating and training a neural network.

@@

The purpose of this paper is to inform readers with no prior knowledge about machine learning. It is, therefore, necessary that we concisely present all of the topics from the beginning. Giving a deeper understanding of the theory is a bigger priority for us than presenting formal mathematical descriptions of everything which could end up making the readers more confused.

@@

The nature of learning is often associated with the foundational idea of learning by interacting with the surrounding environment. When a child learns how to ride a bicycle, the child either falls or manages to stay in balance. Every time parents speak to their children they indirectly correct the children's sentences by repeating them correctly. Constantly interacting with the environment produces a wealth of information about the consequences of actions, their cause and effect of them and what actions to take to achieve a certain goal. These kinds of interactions undoubtedly play a major role throughout our lives as we learn about ourselves and the environment surrounding us. Whether it is learning how to ride a bike or how to speak, we are always aware of how the environment responds to our actions. By behaving in a certain manner and choosing actions carefully, we can reach the goals we seek. Reinforcement learning \cite{sutton} offers a computational approach to learning from interaction. By primarily exploring idealized learning situations and evaluating the effectiveness of various learning methods, reinforcement learning focuses a lot more on goal-oriented learning from interaction than any other machine learning approach.

@@

The following are examples to better grasp the concept of reinforcement learning and how it could be applied to real-life problems.

@@

\begin{outline}
    \1 Adam left his house five minutes ago and is on his way to school. He suddenly starts questioning whether he locked the front door or not. Turning home would at this point make him late to class but leaving the house unlocked has its risks. Adam makes his decision based on the time he gets home after school and how often he has forgotten to lock the door. 
    \1 Hugo is playing chess against the current chess world champion Magnus Carlsen. In order not to lose, Hugo makes moves by evaluating the current position intuitively, and anticipating possible counterreplies. Since Hugo is familiar with the current position and has played it at home several times, he is quite confident in his decisions. 
    \1 Romeo could hardly move his body the first few months after being born. One year later, he was running at 1.6 meters per second.
    \1 Jonathan is in a casino and suddenly sees ten slot machines. He does not know the probability of getting a jackpot on any of the machines but knows the cost of each lever pull and the reward given by a jackpot. Through repeated selections of slot machines, Jonathan manages to maximize his winnings by concentrating on the best levers. \end{outline}

@@

\noindent
All of the examples share features that are easily overlooked due to their simplicity. They all involve interaction between a decision-making agent and an environment, where the agent is trying to achieve a particular goal without knowing the full extent of the environment, nor the consequences of its actions. The agent's actions can affect the environment (e.g., Hugo making a move against Magnus Carlsen changes the state of the board) thereby affecting the actions available to the agent in the future. None of the actions, in all of the examples, can be predicted fully and the agent must instead observe the environment and how it changes based on actions to learn how to properly behave.

@@

As previously illustrated, there are two primary components of reinforcement learning, an agent and an environment. It is, on the other hand, possible to recognize three more elements of a reinforcement learning system. These being a policy, a reward signal, and a value function.

@@

\noindent
Further, the \emph{reward signal} \cite{sutton} determines the goal of the learning agent. After each decision the agent makes, the environment sends a reward signal, and the goal is to maximize the total reward after all of the actions have been made. The reward signal thus determines what actions are good and bad, respectively. This is comparable to pain and pleasure in human beings. When a child falls off their bike, they receive a negative reward signal in the form of pain. On the other hand, maintaining balance on the bike for a certain amount of time sends a positive reward signal in the form of happiness. Since the child, the learning agent, wants to experience joy instead of pain, the goal becomes riding a bike without falling off it.

@@

A Markov Decision Process \cite{sutton}, MDP, is a more rigorous and mathematical way of describing a sequential decision-making system. An MDP can formally be defined as MDP$(S, A, R)$ where each parameter is defined as follows:

@@

\noindent
More specifically, an agent and an environment interact with each other in an MDP at discrete time steps, $t = 0, 1, 2, 3, ...$. The agent receives a representation of the environment's state, $S_{t} \in S$, at each time step and accordingly selects an action, $A_{t} \in A(S)$. The time step following, $t + 1$, as a consequence of its action, the agent receives a reward signal, $R_{t} \in R$, and arrives at a new state, $S_{t+1}$. The MDP thereby gives rise to the following sequence:

@@

\noindent
In our case, with Connect Four, the board becomes the environment, and the players are agents taking turns interacting with it in the Markov Decision Process.

@@

\vskip 0.2cm
In 1989 Watkins \cite{sutton} released one of the early breakthroughs in reinforcement learning. It was the algorithm known as Q-learning which is a way for an agent to learn how to optimally navigate in a Markov Decision Process (3.4). The Q-learning algorithm is an action-value function which maps actions to values, where the action values are called Q-values. The Q-learning algorithm evaluates how beneficial a certain action is in a given situation and a higher Q-value corresponds to a better action. The evaluations made by the algorithm improve over time. The algorithm \cite{sutton} is formally defined as follows:

@@

\noindent
\emph{Gamma}, $\gamma$, is in this expression known as the discount factor and it changes the behaviour in the agent's actions. Values closer to 0 lead to the reward $R_{t}$ propagating worse from the state $S_{t}$ to $S_{0}$ which encourages the agent to take actions that yield immediate rewards. A discount factor closer to the value 1 will on the other hand encourage the agent to take actions that yield later rewards since a larger portion of the reward propagates through the states in this case.

@@

\noindent
\emph{Alpha}, $\alpha$, is in this expression known as the learning rate and regulates the agent's rate of learning, just as the name suggests. Values closer to 0 will decrease the update in the formula. A learning rate closer to 1 will, on the other hand, increase the rate of learning as a large part of the updated value remains.

@@

Since the future maximum action value, $\max_{a}Q(S_{t+1}, a)$, in Q-learning is evaluated using the same Q functions as the one used by the agent to choose the action, $a_{t}$, the algorithm risks overestimating certain action values \cite{doubleq}. In our case with Connect Four, it could be evaluating bad moves that randomly led to a victory the same as good moves that lead to a victory. This slows down the learning since the Q-learning algorithm will later need to re-adjust the overestimated evaluations. A variant called double Q-learning \cite{doubleq} was proposed to correct this. In practice, it uses two separate value functions, $Q_{X}$ and $Q_{Y}$, that are updated in a mutually symmetric fashion using each other. Double Q-learning is defined as follows:

@@

\noindent
Now that the future maximum action is evaluated using a separate value function the overestimation issue has been solved. The double Q-learning algorithm could be seen as two people checking that each other's homework is done correctly which naturally increases the correctness and in turn the effectiveness.

@@

Artificial neural networks \cite{charu} are simulations of the mechanism of learning in biological organisms. The human brain and nervous system contain cells known as neurons. The neurons are connected with the help of synaptic connections. The strength of the synaptic connections changes in response to external stimuli, which is how learning takes place, not only in humans but also in many other living organisms.

@@

\noindent
This learning is simulated in artificial neural networks \cite{charu} which contain artificial neurons. The neurons are connected through weights that imitate the strength of synaptic connections in living organisms.

@@

\noindent
The basic architecture of the perceptron is shown in Figure 10 above. The input layer consists of input nodes that each are connected to the output node through weights, $w_{1}, w_{2}, ..., w_{6}$. The input nodes do not perform any changes to the input values, $x_{1}, x_{2}, ..., x_{6}$ and send the input values to the output node through the corresponding weight connections. The output node then receives the sum of each input value multiplied by the weight it was sent through, a weighted sum. The node also contains an internal value called \emph{bias}, $b$, that is added to the sum. If we let $v$ denote the output node's recieved value, then $v = b + \sum w_{i} x_{i} = w_{1} x_{1} + w_{2} x_{2} + ... + w_{6} x_{6}$. Before the output then sends the value further, it applies an \emph{activation function} which often is non-linear. If we let $f$ denote the activation function, the output of the output node becomes $y = f(v)$. This procedure where a set of input values, $x_{i}$ are transformed into a set of output values, $y_{i}$, by the neural network is called a \emph{forward pass} \cite{charu}.

@@

Multilayer neural networks consist of more than one computational layer \cite{charu} in contrast to the perceptron. The perceptron only consists of an input and output layer, whereas the output layer is the only layer that performs computations. Multilayer neural networks on the other hand contain intermediate layers in addition to the input and output layers. The intermediate layers are between the input and output layers and are often referred to as \emph{hidden layers} since the computations performed by them are not visible. The weight connections, internal biases and the mechanics of the forward pass remain the same as in the perceptron, the only difference being that there are several nodes in each hidden layer which receive a weighted sum of values from the preceding layer.

@@

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.6]{multilayered}
    \vskip 0.2cm
    \caption{The basic architecture of a multilayered neural network with two hidden layers and a single output layer. Note that the input layer is excluded when we count the number of layers since it only transmits the input data into the network. }
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.3]{feedforward}
    \caption{A forward pass in a multilayer neural network with 1 hidden layer and 1 output layer. Weighted sums from the preceding layers are in orange. The value in each node represents its internal bias.}
\end{figure}

\noindent
Figure 12 above illustrates a multilayer neural network with 2 input nodes, followed by a hidden layer with 3 nodes which connects to a single output node. Both the hidden and the output layers use a linear activation function, $f(x) = x$, for simplicity. The two input values are each 1 and the internal bias, $B$, of input nodes is always 0. Let each node be indexed as $(i, j)$ where $i$ is the index of the layer from left to right, zero-indexed, and $j$ is the index of the node in its layer from top to bottom, zero-indexed. Furthermore let $W_{(i_{1}, j_{1}) (i_{2}, j_{2})}$ denote the weight connection between node $(i_{1}, j_{1})$ and $(i_{2}, j_{2})$. The value, $V$, of each node and the final output, $y$, are then calculated as follows:

@@

Figure 12 shows a neural network that only uses a linear activation function, $f(x) = x$, but non-linearity must be introduced into the network. Non-linearity means that the output of the neural network cannot be reproduced from a linear combination of the input values. Without a non-linear activation function in the neural network, no matter how many layers it has, it would behave like a simple perceptron, since combining the layers would only produce another linear function. By introducing non-linear activation functions into the neural network, it can approximate models that vary non-linearly in addition to linear models. The activation function used throughout this project is the hyperbolic tangent function, \emph{tanh}, which has the output range $[-1, 1]$ and is defined as follows:

@@

A neural network can be interpreted as a large statistical model with a large number of parameters, and weight connections. If each of the parameters, weight connections, were to be tuned well, then the model would be able to represent highly abstract and complex correlations. A less rigorous way of imagining this would be having a function in a space with as many dimensions as weight connections in the network. The phenomena that that function would be able to approximate would be beyond our understanding. This leads to the most common usage of neural networks. Namely training neural networks by adjusting their weight connections for it to approximate a certain function with the help of data. An example would be approximating the sinus curve with a neural network. By giving an x-coordinate as input to the network and comparing the network's output value with the y-coordinate one only needs to adjust the weight connections of the network in such a manner that the error decreases to make it a better approximation. This is further described in the following sections.

@@

To be able to train a neural network to a certain level we, first of all, need to know at what level it currently is. An efficient method to measure this is to find an error. Given an output target and the neural network's output, the error can be described as a function of them. To then improve the network we simply need to find a way to decrease that error. This type of function which is to be minimized is also referred to as a \emph{loss function} \cite{charu}. How the loss function is used to train a neural network is described in later sections. The loss function used throughout this project is the \emph{Mean Squared Error} \cite{charu}, MSE, and is defined over a set of network predictions or outputs, $\hat{y_{1}}, \hat{y_{2}}, ..., \hat{y_{n}}$, and output targets, $y_{1}, y_{2}, ..., y_{n}$ as follows:

@@

\noindent
An advantage of The MSE is that it ensures that our trained model will not have outlier predictions with prominent errors, since the MSE puts larger weight on those errors due to the squaring in the function. A disadvantage is, on the other hand, that a single very poor prediction made by the network will overly impact the training since the error is magnified by the squaring part of the MSE function. Since most of the cases aim for more of a well-rounded neural network that performs well on most of the data those outliers should be ignored and not prioritized. This is not a big problem in our project since our networks only output values in the range $[-1, 1]$ which ensures that the squaring in the MSE does not magnify any errors by a lot.

@@

Assume we are trying to train a perceptron with one input node and one output node with one weight connection, $w$, between them. For the sake of simplicity, let the output node's activation function be linear, $f(x) = x$. Let the input value be a constant $x$, the output target be a constant $y$, and the output node's internal bias be a constant $b$. The error is then defined as follows:

@@

\noindent
Notice that $w$ is the only variable in the error and is exactly what needs to be adjusted as mentioned in section 4.3. If we now plot the MSE function as a function of $w$ things become more apparent. The second-degree polynomial created by the MSE in respect to the weight, $w$, is illustrated below, non-scaled:

@@

\noindent
As Figure 15 shows the local minimum of the MSE and finding a weight $w$ such that the loss function is at the local minimum is equivalent to minimizing the error, which is desired as described in section 4.4.1. One way of finding that local minimum is to descend the loss function by using its slope at the current weight, also known as the derivative at that point. Note that we assume that the loss function is defined and differentiable around that point. Since the slope describes the 'steepness' of the loss function in the current point, following the direction of that slope will either lead you to a local maximum or minimum. It could be thought of as walking up or down a hill as illustrated by Figure 15 above. The slope of the MSE changes as one walks down the slope, assuring that the movement is towards the minimum. This type of descent will be described more rigorously in the following sections.

@@

In order to rigorously generalize the descent algorithm described in section 4.4.2 a new mathematical tool needs to be introduced. Namely the \emph{gradient} of a function. The example used in section 4.4.2 was a perceptron with a single output node and a single input node. Having a multilayer neural network with several weight connections instead leads to the loss function being dependant on several variables, namely the weights $w_{1}, w_{2}, ..., w_{n}$. To find the slope of a function $f(w_{1}, w_{2}, ..., w_{n})$ in a point we employ a mathematical technique called \emph{partial derivatives}. Let us use a small example. Take the following simple case of $f$:

@@

\noindent
Using the expression in Figure 21 in combination with the expression in Figure 20 the following recursion is attained:
